---
title: World Happiness Report for 2019
author: Julian Terry S. Bass
output: html_document
---

This is an R Notebook File to support further analysis of the World Happiness project.

Importing Libraries

```{r}

library(ggplot2)
library(plyr)
library(reshape2)
library(knitr)
library(binom) 
library(ggthemes)

```

**Loading in the CSV data for 2019**

```{r}
happiness.data = read.csv(file = 'HappinessData2019.csv', header=TRUE)

# Changing some column names
names(happiness.data)[10] = 'Population.2020'
names(happiness.data)[11] = 'Population.density'
names(happiness.data)[12] = 'Migrants.net'
names(happiness.data)[13] = 'Urban.pop'

```

**Comparing Ratio Variables with Overall Scores**

```{r}

# Subset for Ratio Variables and Scores + melting
scores.ratios = happiness.data[,3:10]
scores.ratios.melt = melt(scores.ratios, measure.vars = names(scores.ratios)[2:7])

# Plotting melted ratio variable data according to population
ggplot(scores.ratios.melt, aes(x = Score, y = value), las = 2) + 
  geom_point(aes(color = value > 0.8)) +
  facet_wrap(~ variable) + 
  geom_jitter(position = position_jitter(width = 0.1, height = 0.1), aes(color = value > 0.8), alpha = 0.5) +
  labs(title = "Ratio Values According to Happiness Score", x='Scores', y='Value by Ratio Variable')


```
The ratio values generally land between 0.2 and 1.5, and we can see that using facet_wrap(), GDP per capita, Social Support and a Healthy Life Expectancy are great factors in determining the happiness score of a nation. This aligns with the line chart previously shown in Power BI. I've used the midpoint of 0.8 to divide the data into a lower and upper half per variable, and we can see the three mentioned above are the only variables that fit into the upper half.

While it may not be as drastic an observation, the freedom to make life choices seems to have a slightly greater influence that perceptions of generosity and corruption.

**Re-doing the plot according to population**

```{r}

# Plotting melted ratio variable data according to population
ggplot(scores.ratios.melt, aes(x = log(Population.2020), y = value), las = 2) + 
  geom_point(aes(color = value > 0.8)) +
  facet_wrap(~ variable) + 
  labs(title = "Ratio Values According to Population", x='Population', y='Value by Ratio Variable') + geom_jitter(position = position_jitter(width = 0.1, height = 0.1), aes(color = value > 0.8), alpha = 0.5)

```
The facet plot above mimics that of the previous plot, but instead of basing everything on the happiness scores, we are now basing it on populations.

The results are mostly the same as with the original plot as expected. Power BI showed us that the population of a nation didn't seem to have any major influence on happiness score ranges, so it's expected that these two facet plots be similar.
Averages for Ratio Variables

**Using geom_violin to look at average values and distributions among variables**

```{r}

variable.averages = ddply(scores.ratios.melt, "variable", summarize, avg.ratio = mean(value))

ggplot(data=scores.ratios.melt, mapping=aes(x = variable, y=value, color=variable)) + 
  geom_violin() + 
  geom_point(position=position_jitter(width=.3), size=.6) + 
  theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1)) + 
  scale_color_discrete(guide=FALSE) + 
  labs(x = 'Variables', y='Values', title='Ratio Variable Values by Melting') +
  stat_summary(fun.y=median, geom="point", size=2, color="red")

ggplot(variable.averages, aes(x=avg.ratio, y=variable)) +
  geom_bar(stat="identity") +
  labs(x = 'Average Values', y='Variables', title='Average Values per Variable')

```
Quoting a reference from https://mode.com/blog/violin-plot-examples/,

"Wider sections of the violin plot represent a higher probability that members of the population will take on the given value; the skinnier sections represent a lower probability."

This violin plot once again shows us that a perception of a high GDP, high levels of Social Support and a Healthy Life Expectancy are the ruling contributors to the decision of a happiness score. While initial readings showed GDP to have the highest values, it may very well be that Social Support actually takes the top spot because it has a higher average value and point distribution. This is further supported by the accompanying barplot of average ratio values.

As mentioned in the reference, wider sections of each violin mean that values will likely be more concentrated on that point. GDP has a generally even distribution while Social Support and Life Expectancy are more concentrated on higher scores. Freedom to make life choices, while being part of the lower trio of variables, shows a noticeably strong concentration in higher values.

**Attempting to create a more uniform labeling system for demographic information with cut_number categories**

```{r}

# Getting complete records
complete.data = happiness.data[ complete.cases(happiness.data), ]

# Converting Urban.pop column percentage to decimal
complete.data$Urban.pop <- as.numeric(sub("%", "",complete.data$Urban.pop,fixed=TRUE))/100

# Categories for Population
complete.data$Population.2020 = cut_number(complete.data$Population.2020, n = 4, labels = c('Class 1 - (Low)', 'Class 2 - (Medium)', 'Class 3 - (High)', 'Class 4 - (Very High)'))

# Categories for Population Density
complete.data$Population.density = cut_number(complete.data$Population.density, n = 4, labels = c('Class 1 - (Low)', 'Class 2 - (Medium)', 'Class 3 - (High)', 'Class 4 - (Very High)'))

# Categories for Migrant Counts
complete.data$Migrants.net = cut_number(complete.data$Migrants.net, n = 4, labels = c('Class 1 - (Low)', 'Class 2 - (Medium)', 'Class 3 - (High)', 'Class 4 - (Very High)'))

# Categories for Urban Population Percentage
complete.data$Urban.pop = cut_number(complete.data$Urban.pop, n = 4, labels = c('Class 1 - (Low)', 'Class 2 - (Medium)', 'Class 3 - (High)', 'Class 4 - (Very High)'))

```
The code above may appear a bit confusing, so we'll break it down into chunks. 

When looking at the original happiness score data, specifically the "demographics" attributes like population, population density, migrant count and degree of urbanization, we can see that the range of values are very far in between. 

For example, the highest rated nation, Finland, has the following demographic attributes:

population: 5540720
population density: 18
net migrants: 14000
urban population: 86%

Because the properties and formats of these data values being so diverse in length, it makes it difficult to plot them on the same graph, so I decided to use the cut_number() function to create a singular range of breaks per attribute. These breaks are:

Class 1 - (Low)
Class 2 - (Medium)
Class 3 - (High)
Class 4 - (Very High)

These classes give a more uniform context to the range of values per attribute. Let's look at the previous record of Finland now with the transformed class values:

population: Class 2 - (Medium)
population density: Class 1 - (Low)
net migrants: Class 3 - (High)
urban population: Class 4 - (Very High)

Personally, I find these much easier to read on a single plot. I did however ask myself if these classes were truly accurate representations of their true numeric values in the original data, that is, does a "Medium" population class really match with the original value of 5540720 for Finland? The answer is yes, they do. I dug through both versions of the data set and compared some records among different nations, and I found that the class representations really did follow through with the original values. Nations with a population class of very high indeed showed values much higher than other nations, and nations with a mid-range or "Medium" migrant count really did have migrant count values generally in the mid-range compared to others.

This means that I have successfully created a uniform classification system that makes the otherwise diverse value range of demographic attributes more readable on a single plot. The plot below shows these classes in action according to the happiness score.

**Plotting from above**

```{r}

# Subset for just the demographic data
scores.demographics = subset(complete.data, 
                             select=c('Country.or.region', 'Score', 'Population.2020', 'Population.density', 'Migrants.net', 'Urban.pop'))

# Melting into one column
scores.demographics.melt = melt(scores.demographics, measure.vars = names(scores.demographics)[3:6])

scores.demographics.melt = scores.demographics.melt[ complete.cases(scores.demographics.melt), ]

ggplot(scores.demographics.melt, aes(Score, value, color = value)) + 
  geom_point(position=position_jitter(width=.3), size=.6) + 
  stat_smooth() +
  scale_x_continuous(breaks = c(2, 3, 4, 5, 6, 7, 8)) +
  facet_wrap(~ variable, nrow = 4) +
  labs(title='Demographic Data by Happiness Score')

```

