---
title: ANLYTC3 Homework 4 
author: Julian Terry S. Bass
output: html_document
---

#### Instructions 

The due date is 09:30AM on Feb 23 (Tue). Submit BOTH the Knit HTML file and the Rmd file.

Code should be clearly commented. Plots should be presentable and properly labeled. Mitigate overplotting whenever possible.

#### Preliminaries

We'll use the data file `dataset_hw4.csv`, which should be in the same directory as this markdown file (which should also be your working directory). It is a data frame of expenditures by household from the consumer expenditure survey


#### Questions 


**Problem 1:** 

Group the households into two categories:

* apartments: those whose `home.type` is "apartment or flat" or "high-rise"
* homes: those whose `home.type` is "single family detached"

Compare the `housing` expenditures for the two groups using a quantile plot. Do they appear different? 


```{r fig.width=8, fig.height=6, dpi=100, fig.align='center'}
library(plyr)
library(ggplot2)
library(reshape2)
expenditures = read.csv(file = 'dataset_hw4.csv', header = TRUE)
```

```{r fig.width=8, fig.height=6, dpi=100, fig.align='center'}
# fill in

# Subset for home categories
apartments.and.homes = subset(expenditures, home.type == "apartment or flat" | home.type == "high-rise" | home.type == "single family detached")


qplot(sample = housing, data = apartments.and.homes, color=home.type) + facet_wrap(~ home.type)

```

ANS: 

**Problem 2:**

Compare the `housing` expenditures but first take a log transform. (Exclude those who spent zero or negative amounts on `housing` from the analysis)

Use quantile plots and/or QQ plots to compare the two groups after the log transform. If helpful, also compare the residuals. What are the similarities and differences between the two groups?

```{r fig.width=8, fig.height=6, dpi=100, fig.align='center'}
# fill in

qplot(sample = housing, data = apartments.and.homes, color=home.type) + 
  facet_wrap(~ (housing > 5000) + home.type)

```

ANS: 

** Problem 3:**

Using the `movies` data set that comes with `ggplot2`, visualize the distribution of ratings for movies before and after 1930.

How do the distributions differ? Should the distributions be compared with or without a log transform?

You may want to adjust the size of the points to reduce overplotting.

```{r fig.width=8, fig.height=6, dpi=100, fig.align='center'}
# fill in
library(ggplot2movies)

ggplot(movies, aes(x = year, y = rating)) +
    geom_line(aes(color = factor(year > 1930))) +
    stat_summary(fun.data = mean_cl_boot, geom = "point", color = "black")

```

ANS: 

** Problem 4:**

Compare the distribution of ratings for each of the genres: action, animation, comedy, drama, documentary, romance, and short. If a movie belongs to more than one genre, include it in both distributions. Use both quantile plots and Q-Q plots.

```{r fig.width=8, fig.height=6, dpi=100, fig.align='center'}

# fill in

# Re-assigning movies to remove null entries
movies = na.omit(movies)
# Repeat function repeats as many as the total rows in movies
genre = rep(NA, nrow(movies))
# Getting the sum of the binary row values for each movie across the different genres
# 18:24 represents the columns for the genres
# For context, if the sum is greater than 1, the movie is part of multiple genres
count = rowSums(movies[, 18:24])

# Conditional checking for each record
# count == 1 means a movie is part of one genre
genre[which(count == 1 & movies$Action == 1)] = "Action"
genre[which(count == 1 & movies$Animation == 1)] = "Animation"
genre[which(count == 1 & movies$Comedy == 1)] = "Comedy"
genre[which(count == 1 & movies$Drama == 1)] = "Drama"
genre[which(count == 1 & movies$Documentary == 1)] = "Documentary"
genre[which(count == 1 & movies$Romance == 1)] = "Romance"
genre[which(count == 1 & movies$Short == 1)] = "Short"
movies$genre = as.factor(genre)

ggplot(movies, mapping = aes(x = year, y = rating, color = rating)) + 
  stat_summary(fun = mean, geom = "line", size = 0.8) +
  geom_smooth(color = "red") +
  facet_wrap(~ genre)
```

True or False: 
* The ratings distribution for action movies is worse than those of the pooled movies
* Animation movies have better ratings than the overall distribution at the lower quantiles, but worse than the overall distribution at the highest quantiles.
* Documentaries and Shorts have worse ratings than the overall distribution at the lower quantiles, but better ratings than the overall distribution otherwise.

Also, which worked better for answering the T/F questions: quantile plots or QQ plots?

ANS: 

** Problem 5:**

Compare the distribution of ratings for each of the genres, and also for the following time periods: 1900-1920, 1921-1940, 1941-1960, 1961-1980, 1981-2000 (i.e., there should be 35 groups total). Use both quantile plots and Q-Q plots. 

```{r fig.width=10, fig.height=10, dpi=100, fig.align='center'}
# fill in
```

* Describe how the distribution of ratings for action movies changed over time, relative to the pooled ratings
* Describe how the distribution of ratings for comedies changed over time, relative to the pooled ratings

ANS:

